---
title: "Assignment 2"
author: "Emmanuella Ekpenyong (2696204), Fie Bergli (2902784) & Britta Abbink Spaink (2772949)"
date: "10 March 2026"
output: pdf_document
fontsize: 11pt
highlight: tango
---

## Exercise 1
### 1a) 

```{r}
birth_weight <- read.csv("Birthweight.csv", header=TRUE)
```

To test the relationship between smoking and low birth weight we define a null hypothesis (H0): smoking and low birth weight are independent. We also define an alternative hypothesis (H1): smoking and low birth weight are associated. To test these hypothesis', we apply a contingency table.  
```{r}
table_smoke <- table(birth_weight$smoker, birth_weight$lowbwt)
```

```{r}
fisher.test(table_smoke)
```
A Fisher's Exact Test was used instead of the Chi-square test because some expected counts in the contingency table were below 5, making the Chi-square approximation unreliable. The Fisher's test gives a p-value of 0.187. As our confidence level is a = 0.05, we see that 0.187 > 0.05. This result fails to reject the null hypothesis, which means there is no statistically significant relationship between maternal smoking and low birth weight in this data set. 
Interestingly, the odds ratio suggest a possible increase in the odds of low birth weight among smokers. However, the wide confidence interval indicates substantial uncertainty, likely due to the small sample size. 

To test the relationship between the mother having an age of 35 or more and low birth weight we define a null hypothesis (H0): maternal age is 35 or more and low birth weight are independent. We also define an alternative hypothesis (H1): maternal age is 35 or more and low birth weight are associated. 
```{r}
table_mage35 <- table(birth_weight$mage35, birth_weight$lowbwt)
```

```{r}
fisher.test(table_mage35)
```
The Fisher's Exact Test produced a p-value of 0.4737. Since the given significance level is a = 0.05, we observe that 0.4737 > 0.05. Therefore, we fail to reject the null hypothesis, which means there is no statistically significant relationship evidence of an association between the mother being aged 35 or older and low birth weight in this data set. 

##1b)
```{r}
library(glmnet)
set.seed(1)
```

```{r}
y <- birth_weight$Birthweight
X <- model.matrix(Birthweight ~ ., birth_weight)[, -1]
```
```{r}
cv_lasso <- cv.glmnet(X, y, alpha = 1)
best_lambda <- cv_lasso$lambda.1se
best_lambda
```
```{r}
lasso_model <- glmnet(X, y, aplha=1, lambda = best_lambda)
coef(lasso_model)
```
```{r}
full_model <- lm(Birthweight ~ ., data = birth_weight)
summary(full_model)
```
```{r}
pred_full <- predict(full_model, newdata = birth_weight)
pred_lasso <- predict(lasso_model, newx = X)

mse_full <- mean((y-pred_full)^2)
mse_lasso <- mean((y-pred_lasso)^2)

mse_full
mse_lasso
```

The lasso model selected a subset of predictors by shrinking several coefficients to zero. The performance of the lasso model prediction was compared to a full linear regression model using the mean squared error. The full model achieved a lower MSE of 0.069, compared to the lasso model that achieved a MSE of 0.112. This indicates that the full model has better prediction accuracy on this data set, although the lasso provides a simpler model by selecting only the most relevant predictors.


## Exercise 2 
## 2a) Study the data by giving a few summaries (graphical or tables), make preliminary comments related to the main question.
```{r}
custody <- read.table("custody.txt", header = TRUE)

custody$Indigenous <- as.factor(custody$Indigenous)

summary(custody)

aggregate(cbind(Deaths, Prisoners) ~ Indigenous, data = custody, sum)
```
```{r}
custody$DeathRate <- custody$Deaths / custody$Prisoners
aggregate(DeathRate ~ Year + Indigenous, data = custody, mean)
aggregate(DeathRate ~ Indigenous, data = custody, mean)
```

```{r}
indig <- subset(custody, Indigenous == "Yes")
nonindig <- subset(custody, Indigenous == "No")

plot(indig$Year, indig$DeathRate,
     type = "o", pch = 19, col = "red",
     ylim = range(custody$DeathRate),
     xlab = "Year", ylab = "Death rate")

lines(nonindig$Year, nonindig$DeathRate,
      type = "o", pch = 19, col = "blue")

legend("topleft",
       legend = c("Indigenous", "Non-Indigenous"),
       col = c("red", "blue"),
       pch = 19,
       lty = 1)
```
The number of prisoners increases slightly from 1990 to 1995 for both Indigenous and non-Indigenous groups, although the Indigenous prison population is much smaller. The number of deaths fluctuates across years, with higher raw counts for non-Indigenous prisoners due to their much larger prison population. A more informative measure is the death rate (Deaths/Prisoners). The exploratory analysis suggests that the death rate is generally higher for Indigenous prisoners, although it varies considerably from year to year. These observations motivate a formal statistical analysis using logistic regression to determine whether the difference between the groups is statistically significant.

## 2b) To address the main question, fit a logistic regression model by modelling the number of deaths (among prisoners) as response, Year as a continues predictor and Indigenous as a factor predictor, for now insisting on presence of both predictors. Is there an interaction between Year and Indigenous?

```{r}
model1 <- glm(cbind(Deaths, Prisoners - Deaths) ~ Year + Indigenous,
              data = custody,
              family = binomial)

summary(model1)
```

```{r}
model2 <- glm(cbind(Deaths, Prisoners - Deaths) ~ Year * Indigenous,
              data = custody,
              family = binomial)

anova(model1, model2, test = "Chisq")
```
A logistic regression model was fitted with the number of deaths among prisoners as the response variable, with Year as a continuous predictor and Indigenous status as a factor predictor. An additional model including the interaction between Year and Indigenous was fitted to test whether the time trend differs between the groups. A likelihood ratio test comparing the models gave a p-value of 0.2617. Since this is greater than 0.05, the interaction is not statistically significant. Therefore, there is no evidence that the effect of Year differs between Indigenous and non-Indigenous prisoners.

## 2c) For the model from b), estimate the relative change in the odds (odds ratio) of death in custody from non-indigenous to indigenous group. What is the interpretation of this number? Does it depend on year? Demonstrate the eﬀect of Year and Indigenous by plotting the odds from 1990 to 1995 for both groups, Indigenous and non-Indigenous, on the same plot. Comment.
```{r}
exp(coef(model1)["IndigenousYes"])
```
The estimated odds ratio is 1.29, indicating that the odds of death in custody for Indigenous prisoners are approximately 29% higher than for non-Indigenous prisoners, holding Year constant. Since the interaction between Year and Indigenous was not statistically significant, the odds ratio does not depend on year. Thus, the relative difference between the two groups remains constant over the observed time period.
```{r}
years <- 1990:1995

newdata_yes <- data.frame(Year = years, Indigenous = "Yes")
newdata_no  <- data.frame(Year = years, Indigenous = "No")

p_yes <- predict(model1, newdata_yes, type = "response")
p_no  <- predict(model1, newdata_no, type = "response")

odds_yes <- p_yes/(1-p_yes)
odds_no  <- p_no/(1-p_no)

plot(years, odds_yes, type="o", col="red", pch=19,
     ylim=range(c(odds_yes, odds_no)),
     xlab="Year", ylab="Odds of death in custody")

lines(years, odds_no, type="o", col="blue", pch=19)

legend("topleft",
       legend=c("Indigenous","Non-Indigenous"),
       col=c("red","blue"),
       pch=19,
       lty=1)
```
The predicted odds from 1990 to 1995 were plotted for both groups, showing higher odds for Indigenous prisoners across all years. The difference between the groups remains approximately constant over time.


## 2d) Try to extend the model in b) by including Population (use at most one interaction term per model). Is it a good idea to include the variable Population? Using all the results obtained so far, choose (and justify) a resulting model. Comment on your findings.

```{r}
model_pop <- glm(cbind(Deaths, Prisoners - Deaths) ~ Year + Indigenous + Population,
                 data = custody,
                 family = binomial)

anova(model1, model_pop, test="Chisq")
```
```{r}
model_pop_int <- glm(cbind(Deaths, Prisoners - Deaths) ~ Year + Indigenous + Population + Indigenous:Population,
                     data = custody,
                     family = binomial)

anova(model_pop, model_pop_int, test="Chisq")
```
The model from part (b) was extended by including Population as an additional predictor. A likelihood ratio test comparing the models with and without Population gave a p-value of 0.2524, indicating that Population does not significantly improve the model. An additional model including the interaction between Indigenous and Population was also considered, but this interaction was not statistically significant (p = 0.8988). Since neither Population nor the interaction term improved the model, their inclusion is not justified. Therefore, the final model includes only Year and Indigenous as predictors, which is the simplest model not contradicted by the data.

## 2e) Now investigate whether the predictors Year and Indigenous have eﬀect on the proportion of prisoners (relative to the Population). Should the interaction be used in the model?
```{r}
model_prison <- glm(cbind(Prisoners, Population - Prisoners) ~ Year + Indigenous,
                    data = custody,
                    family = binomial)

model_prison_int <- glm(cbind(Prisoners, Population - Prisoners) ~ Year * Indigenous,
                        data = custody,
                        family = binomial)

anova(model_prison, model_prison_int, test="Chisq")
```
To investigate whether Year and Indigenous status affect the proportion of prisoners relative to the population, a logistic regression model was fitted with the number of prisoners among the population as the response variable and Year and Indigenous as predictors. A second model including the interaction between Year and Indigenous was also fitted. A likelihood ratio test comparing the models gave a p-value of 4.418e-08, indicating that the interaction term is highly significant. This suggests that the change over time in the proportion of prisoners differs between Indigenous and non-Indigenous populations. Therefore, the interaction between Year and Indigenous should be included in the model.

## 2f) Combine all your findings so far to present your final conclusion(s) for the Royal Commission on Aboriginal Deaths in Custody.
The exploratory analysis suggested that the death rate among prisoners is generally higher for Indigenous prisoners than for non-Indigenous prisoners, although the rates vary between years. Logistic regression modelling showed that the odds of death in custody for Indigenous prisoners are approximately 1.29 times the odds for non-Indigenous prisoners, holding Year constant. The interaction between Year and Indigenous was not statistically significant, indicating that the relative difference between the groups remains approximately constant over time. Including Population in the model did not significantly improve the model fit, so the simpler model with Year and Indigenous was retained. However, modelling the proportion of prisoners relative to the population showed a significant interaction between Year and Indigenous, indicating that the change over time in imprisonment rates differs between the two groups. Overall, the results suggest that Indigenous prisoners have somewhat higher odds of death in custody, while a much larger disparity exists in the proportion of Indigenous people imprisoned relative to their population size.





